Garaphlet Analysis for HiC data

### 2018-03-15 papers/manuscripts/manuscript.tex
wrote manuscript

Implementations of algorithms of extracting graphlets: \\
\begin{itemize}
    \item GraphCrunch: 
        \url{http://www0.cs.ucl.ac.uk/staff/natasa/graphcrunch2/usage.html}
    \item PGD: \url{http://nesreenahmed.com/graphlets/}
    \item ORCA: Graphlet and orbit counting algorithm \\
        \url{ https://CRAN.R-project.org/package=orca} \\
        This package is in R. In order to install it, type
        \texttt{install.packages("orca")}.

\end{itemize}

The authors of \cite{di2010fast} generalized the idea of graphlets to 
ordered graphs were the nodes are labeled in ascending order.
These graphlets are illustrated in Figure \ref{fig:ordered_graphlets}.
As can be viewed, there are a total of 14 orbits for graphlets of size
2 and 3 since the label of graphlets is also included in toplogy.
In the new definition, $d_v^i$ denotes the number of orbit $i$ touches 
node $v$. Each node, is then assigned a vector of length 14 
\footnote{number of orbits in graphlets of size 2 and 3}
$(d_v^1, d_v^2, ..., d_v^{14})$ 
and similarity of two nodes in two contact maps can be compared by
how geometrically close their corresponding vectors are.
\section{Materials and Methods}
\subsection{Thresholding contact maps}
In order to be able to extract graphlets, HiC contact maps should be modeled as
unweighted graphs where the nodes represent the loci and an edge between two 
nodes represent a \textit{significant} interaction between the loci the nodes
represent.

@@

In order to threshold the matrix so that both global and local patterns are
kept, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity that is larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by an sliding kernel 
that passes through the contact map with a pixel at its middle at 
each step.

\section{Resources}
\textbf{Publications related to Hi-C:}
\begin{enumerate}
    \item \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2858594/}
    \item \url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0058793 }
    \item \url{http://nar.oxfordjournals.org/content/42/7/e52.full}
    \item \url{http://bioinformatics.oxfordjournals.org/content/early/2015/12/31/bioinformatics.btv754.abstract?keytype=ref&ijkey=A97WhKqBiEIcuzd}
    \item \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4417147/}
    \item \url{http://www.pnas.org/content/112/47/E6456.full}
    \item \url{http://www.pnas.org/content/113/12/E1663.full}
\end{enumerate}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item Original Datasets: \url{https://bcm.app.box.com/v/aidenlab/folder/11234760671}
    \item Including cancerous cells: \url{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
    \item Chromosome3D project: \url{http://sysbio.rnet.missouri.edu/bdm_download/chromosome3d/}
\end{enumerate}
\textbf{Contact Matrix Analysis:}
\begin{enumerate}
    \item \url{https://omictools.com/contact-matrix-normalization-category}
    \item \url{http://hifive.docs.taylorlab.org/en/latest/}
\end{enumerate}

\textbf{Labs working on 3D Human Genome:}

\begin{enumerate}
    \item \url{http://mirnylab.mit.edu}
    \item \url{http://dostielab.biochem.mcgill.ca}
    \item \url{http://www.aidenlab.org/}
    \item \url{http://web.cmb.usc.edu/people/alber/index.htm}
    \item \url{http://calla.rnet.missouri.edu/cheng/nsf_career.html}
\end{enumerate}
\textbf{Resources related to Graphlet:}

\begin{enumerate}
    \item \url{https://en.wikipedia.org/wiki/Graphlets}
    \item \url{https://academic.oup.com/bioinformatics/article/23/2/e177/202080/Biological-network-comparison-using-graphlet}
    \item \url{http://www0.cs.ucl.ac.uk/staff/N.Przulj/index.html}
\end{enumerate}

>>>>

The authors of \cite{di2010fast} generalized the idea of graphlets to 
ordered graphs were the nodes are labeled in ascending order.
As can be viewed, there are a total of 14 orbits for graphlets of size
2 and 3 since the label of graphlets is also included in toplogy.
In the new definition, $d_v^i$ denotes the number of orbit $i$ touches 
node $v$. Each node, is then assigned a vector of length 14 
\footnote{number of orbits in graphlets of size 2 and 3}
$(d_v^1, d_v^2, ..., d_v^{14})$ 
and similarity of two nodes in two contact maps can be compared by
how geometrically close their corresponding vectors are.
\section{Materials and Methods}
\subsection{Thresholding contact maps}
In order to be able to extract graphlets, HiC contact maps should be modeled as
unweighted graphs where the nodes represent the loci and an edge between two 
nodes represent a \textit{significant} interaction between the loci the nodes
represent.

@@

In order to threshold the matrix so that both global and local patterns are
kept, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity that is larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by an sliding kernel 
that passes through the contact map with a pixel at its middle at 
each step. Figure \ref{local_thresholded_chr1_chr1} demonstrates result of 
this thresholding approach for interchromosomal contact map of chromosome 1.
Refer to supplementary material for all 23 interchromosomal thresholding
results.
\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/local_thresholded_chr1_chr1.png}
    \caption{Result of thresholding interchromosomal contact map of chromosome 1
    using a kernels of size $5 \times 5$. The first row shows the thresholded
    maps. Second and third rows demonstrate pair-wisesimilarities and 
    differences between contact maps respectively.}
    \label{local_thresholded_chr1_chr1}
\end{figure*}

\subsection{Orbit Extraction}
Once the thresholded contact maps are obtained, graphlets and orbits can be 
extracted. We used the \texttt{orca} package in \texttt{R} programming 
language to extract the graphlets. As a result of graphlet extraction, 
For each loci in each contact map, a \textit{signature vector} of size
73 is created.

We divied the task of graphlet comparison into two parts: first we compare
graphlets from each contact map in normal cell lines (MIT) with the 
same contact maps in the other three Leukemic cells. Second we compare 
contact maps in a similar way but this time only between leukemic cells.
In the former case, the null hypothesis is that there is no difference
between contact maps of normal cells and leukemic celss and in the later
case, the null hypotheis is that there is no difference between 
different leukemic cells.

We consider two measures of \textit{difference} when comparing contact
map graphlets. The first measure is the distance between signature vectors
of a loci in two cell lines. If $S^A_{k}$  and $S^B_{k}$ are 
signature vectors corresponding to loci $k$ in contact maps A and B 
respectively, then we can caclulate the distance between the two loci 
using the following formula from \cite{prvzulj2007biological}:

\begin{equation}
    d_{ki} = w_i \times \frac{log(S^A_{ki}+1) - log(S^B_{ki}+1)}{log(max(S^A_{ki}, S^B_{ki}) + 2)}
\end{equation}
The distance between $S$ and $S'$ can be calculated as follows:
\begin{equation}
    D(S^A_k, S^B_k) = \sqrt{\sum_{0}^{72}{d_{ki}^2}}
\end{equation}

Using this distance measure, we can quantify how two loci are close to
each other in terms of local neighborhood between the two contact maps.

The second measure of comparison that we use captures how 
similar two orbits are in terms of their frequencies across loci
between the two contact maps. For each contact map, we can calculate
similarity between orbit distributions by calculating pearso's r 
correlation. However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning that the two variables are completely independent
and 1 meaning they are totally correlated.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

\subsection{Hypotheses}
\subsubsection{terminology}
We denote the set of all contact maps of normal (MIT) cell line with
$CM$. In the same manner we denote $CA$, $CR$ and $CC$ as the set of all
contact maps corresponding to ALL, RL and CALL4 cell lines.
We denote a member of any of the aforementioned sets using small
letters of the set name. For example, we denote any member of $CA$ with
$ca$.

We denote a contact map by $C^{\{cell_{\{chr1\}\{chr2\}}\}}$, where the
superscript denotes the cell type and the subscripts in the superscripts
determines the pair of chromosomes the contact map is about. For 
example $C^{MIT_{1,14}}$ denotes the contact map between chromosomes
1 and 14 in MIT (Normal) dataset. In such a manner $C^{MIT_{1,14}}_{ij}$
denotes the interaction between loci $i$ and $j$ in the aforementioned
contact map.

We denote $S^C_i$ as the extracted 
signature vector of locus $i$ in contact map $C$. 
Also $P^C_k$ denotes
the frequency distribution of orbital $k$ in contact map $C$.

With regard to the two aforementioned comparison methods, we define
our two null hypotheses as the following:

\begin{enumerate}
    \item For a locus in a contact map in MIT cell type, its distance
        from the same locus in the same contact map of the other three
        cell types are equal to $0$.
            \begin{align}
                \begin{split}
                    H_0: D(sm, sa) = 0  \\
                \hspace{.5cm} \forall cm \in CM \quad\&\quad ca \in CA
                \end{split} \\
                \begin{split}
                    H_0: D(sm, sr) = 0   \\
                \hspace{.5cm} \forall cm \in CM\quad \&\quad cr \in CR
                \end{split} \\
                \begin{split}
                    H_0: D(sm, sc) = 0   \\
                \hspace{.5cm} \forall cm \in CM\quad \&\quad cc \in CC
                \end{split}
            \end{align}
    \item For an orbit extracted from a contact map of MIT cell, the 
        Pearson's r correlation between its frequency and the frequency
        of the same orbit extracted from the same contact map in 
        the other three cell types in equal to 1.
        \begin{align}
            H_0: Corr(P^CM_k, P^CA_k) = 1 \\
            H_0: Corr(P^CM_k, P^CR_k) = 1 \\
            H_0: Corr(P^CM_k, P^CC_k) = 1
        \end{align}

        or

        For an orbit extracted from a contact map of MIT cell, the 
        MIC between its frequency and the frequency
        of the same orbit extracted from the same contact map in 
        the other three cell types in equal to 1.
\end{enumerate}

\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}
